---
title: "UFP predictions from the original Saha et al. 2021 Model"
author: "Magali Blanco"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Notes

* UFP predictions have been updated       
  * the new predictions use recreated (similar) 1k grid UFP data to fit the model     
  * we generate predictions for all blocks (not just those with pop>0 or those within the geocovariate modeling space)        
  * predictions are for all 2010 Census blocks (like original model) as well as 2020 Census blocks 
  * to address extreme UFP predictions (e.g., Manhattan), we winsorize the predictions below/above the 2nd and 98th quantiles 
    *
    * there are many ways to do this (e.g., could winsorize the covariates). This is probably the most straightforward since we are 


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, 
                      cache=F, cache.comments = F, 
                      message = F, warning = F, 
                      tidy.opts=list(width.cutoff=60), tidy=TRUE )  

# Clear workspace of all objects and unload all extra (non-base) packages
rm(list = ls(all = TRUE))
if (!is.null(sessionInfo()$otherPkgs)) {
  res <- suppressWarnings(
    lapply(paste('package:', names(sessionInfo()$otherPkgs), sep=""),
           detach, character.only=TRUE, unload=TRUE, force=TRUE))
}

pacman::p_load(tidyverse, sf,
               knitr, kableExtra
               )  

# ggplot settings
theme_set(theme_bw())
theme_update(legend.position = "bottom")

set.seed(1)
##################################################################################################
# speed thigns up
testing_mode <- TRUE #TRUE #e.g., reduce visit designs & windows/quantile combinations

##################################################################################################
# VARIABLES
##################################################################################################
project_crs <- 4269 #lat/long CRS commonly used in US 


##################################################################################################
# LOAD DATA
##################################################################################################
# quantile summary files
ufp_quantiles_2010 <- readRDS(file.path("output", "winsorize", "raw_quantiles_ufp_covs_2010.rds")) %>%
  mutate(year=2010)

ufp_quantiles_2020 <- readRDS(file.path("output", "winsorize", "raw_quantiles_ufp_covs_2020.rds")) %>%
  mutate(year=2020) 


# raw UFP predictions
#ufp_2010_original <- readRDS(file.path("output", "predictions", "original_model", "blocks", "raw", "ufp_covs_2010.rds"))
# ufp_2020_original <- readRDS(file.path("output", "predictions", "original_model", "blocks", "raw", "ufp_covs_2020.rds"))

# winsorized predictions
ufp_2010 <- readRDS(file.path("output", "predictions", "original_model", "blocks", "modified", "ufp_covs_2010.rds")) %>%
  rename(ufp_original=ufp) %>%
  mutate(year=2010)

if(testing_mode==TRUE){
  ufp_2010 <- slice_sample(ufp_2010, n = 1e3)
  }

# ufp_2020 <- readRDS(file.path("output", "predictions", "original_model", "blocks", "modified", "ufp_covs_2020.rds"))  %>%
#   rename(ufp_original_ufp=ufp) %>%
#   mutate(year=2010)




```

# --> add census block? 

```{r}
# study_counties <- c("King County", "Snohomish County", "Pierce County", "Kitsap County", "Island County")
# 
# tract_path <- file.path("data", "raw", "wa_kc_census_tracts.rda")
# if(file.exists(tract_path)) {
#   tracts0 <- readRDS(tract_path)
# } else {
#   tracts0 <- tracts(state = "WA", county = study_counties, year=yr) %>%
#     filter(ALAND!=0) %>%
#     st_transform(crs = lat_long_crs) 
#   saveRDS(tracts0, tract_path)
# }


```





# quantile summary

--> currently winsorizing at Q 2% and 98% 

```{r}
bind_rows(ufp_quantiles_2010, ufp_quantiles_2020) %>%
  filter(quantile %in% c(0.02,0.98,
                         0.01, 0.99
                         )) %>%
  select(year, everything()) %>%
  mutate(conc = round(conc),
         year = as.character(year)) %>%
  kable(caption = "Quantiles used to winsorize year-specific predictions",
        format.args = list(big.mark=",")) %>%
  kable_styling()


```



# compare original & winsorized UFP predictions 

histogram 

```{r}
ufp_2010 %>%
  select(block_key, contains("ufp")) %>%
  pivot_longer(cols=contains("ufp")) %>%
  
  ggplot(aes(x=value, fill=name)) + 
  geom_histogram(position = "dodge")

# # scatterplot
# ufp_2010 %>%
#   ggplot(aes(x=ufp_original, y=ufp_winsorized, col=winsorized)) + 
#   geom_abline(slope = 1, intercept = 0, linetype=2, alpha=0.5) +
#   geom_point(alpha=0.1)  

```

map of winsorized vs non-winsorized

# --> ERROR: lat/long are wrong 

```{r}
ufp_2010 %>%
  st_as_sf(coords = c('long_block', 'lat_block'), crs=project_crs) %>%
  
  ggplot(aes(col=winsorized)) + 
  geom_sf()



```




