---
title: "UFP predictions from the original Saha et al. 2021 Model"
author: "Magali Blanco"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Notes

* UFP predictions have been updated       
  * the new predictions use recreated (similar) 1k grid UFP data to fit the model     
  * we generate predictions for all blocks (not just those with pop>0 or those within the geocovariate modeling space)        
  * predictions are for all 2010 Census blocks (like original model) as well as 2020 Census blocks 
  * to address extreme UFP predictions (e.g., Manhattan), we winsorize the predictions below/above the 2nd and 98th quantiles 
    *
    * there are many ways to do this (e.g., could winsorize the covariates). This is probably the most straightforward since we are 


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, 
                      cache=F, cache.comments = F, 
                      message = F, warning = F, 
                      tidy.opts=list(width.cutoff=60), tidy=TRUE )  

# Clear workspace of all objects and unload all extra (non-base) packages
rm(list = ls(all = TRUE))
if (!is.null(sessionInfo()$otherPkgs)) {
  res <- suppressWarnings(
    lapply(paste('package:', names(sessionInfo()$otherPkgs), sep=""),
           detach, character.only=TRUE, unload=TRUE, force=TRUE))
}

pacman::p_load(tidyverse, sf,
               knitr, kableExtra,
               tigris
               )  

# ggplot settings
theme_set(theme_bw())
theme_update(legend.position = "bottom")

set.seed(1)
##################################################################################################
# speed thigns up
testing_mode <- FALSE #TRUE #e.g., reduce visit designs & windows/quantile combinations

##################################################################################################
# VARIABLES
##################################################################################################
yr <- 2020
project_crs <- 4269 #lat/long CRS commonly used in US 


##################################################################################################
# LOAD DATA
##################################################################################################
# quantile summary files
ufp_quantiles_2010 <- readRDS(file.path("output", "winsorize", "raw_quantiles_ufp_covs_2010.rds")) %>%
  mutate(year=2010)

# ufp_quantiles_2020 <- readRDS(file.path("output", "winsorize", "raw_quantiles_ufp_covs_2020.rds")) %>%
#   mutate(year=2020) 

# winsorized predictions
ufp_2010 <- readRDS(file.path("output", "predictions", "original_model", "blocks", "modified", "ufp_covs_2010.rds")) %>%
  rename(ufp_original=ufp) %>%
  mutate(year=2010)

if(testing_mode==TRUE){
  ufp_2010 <- slice_sample(ufp_2010, n = 1e3)
  }

# ufp_2020 <- readRDS(file.path("output", "predictions", "original_model", "blocks", "modified", "ufp_covs_2020.rds"))  %>%
#   rename(ufp_original_ufp=ufp) %>%
#   mutate(year=2010)

##################################################################################################
# census data
## 2013-2022 available
state_file <- file.path("output", "other", "states.rda")
if(file.exists(state_file)) {
  states <- readRDS(state_file)
  } else{
    states <- states(year = yr, cb = T #crude resolution is OK
             ) %>%
      filter(!NAME %in% c("Alaska", "Hawaii", "Puerto Rico", "Commonwealth of the Northern Mariana Islands", "Guam", "United States Virgin Islands", "District of Columbia", "American Samoa"))

    saveRDS(states, state_file)
  }


```

# missing predictions

these blocks have missing predictions due to missing model covariates 

```{r}
ufp_2010 %>%
  mutate(missing_imp_a00750 = ifelse(is.na(imp_a00750), "missing imp_a00750", "UFP prediction available" )) %>%
  filter(missing_imp_a00750 == "missing imp_a00750") %>%
  st_as_sf(coords = c('long_block', 'lat_block'), crs=project_crs) %>%
  
  ggplot() + 
  geom_sf(data = states, fill=alpha(0.01)) +
  geom_sf(aes(col=missing_imp_a00750), alpha=1)

```


# quantile summary

Currently winsorizing at the 1st and 99th percentiles 

```{r}
bind_rows(ufp_quantiles_2010#, 
          #ufp_quantiles_2020
          ) %>%
  filter(quantile %in% c(0.02,0.98,
                         0.01, 0.99
                         )) %>%
  select(year, everything()) %>%
  mutate(conc = round(conc),
         year = as.character(year)) %>%
  kable(caption = "Quantiles used to winsorize year-specific predictions",
        format.args = list(big.mark=",")) %>%
  kable_styling()


```



# compare original & winsorized UFP predictions 

histogram 

```{r}
ufp_2010 %>%
  select(block_key, contains("ufp")) %>%
  pivot_longer(cols=contains("ufp")) %>%
  
  ggplot(aes(x=value, fill=name)) + 
  geom_histogram(position = "dodge") + 
  labs(title = "Distribution of original and winsorized UFP predictions")

# scatterplot
ufp_2010 %>%
  ggplot(aes(x=ufp_original, y=ufp_winsorized, col=winsorized)) +
  geom_abline(slope = 1, intercept = 0, linetype=2, alpha=0.5) +
  geom_point(alpha=0.1) + 
  labs(title = "Comparison of UFP predictions before and after winsorizing ")

```

map of winsorized vs non-winsorized

```{r}
ufp_2010 %>%
  st_as_sf(coords = c('long_block', 'lat_block'), crs=project_crs) %>%
  
  ggplot() + 
  geom_sf(data = states, fill=alpha(0.01)) +
  geom_sf(aes(col=winsorized), alpha=0.4) + 
  labs(title = "Map of original and winsorized UFP predictions")



```




